Packing limits and developmental integration are not sufficient to explain stomatal anatomical evolution in flowering plants

### 2022-09-11 ms/ms.Rmd
minor revisions

---
output:
  bookdown::pdf_document2:
    includes:
      before_body: ../template/amnat_doc_prefix.tex
      in_header: ../template/amnat_preamble.tex
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: no
    toc: no
  bookdown::html_document2:
    number_sections: no
    theme: readable
    toc: yes
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
  bookdown::word_document2: null
fontsize: 12pt
linestretch: 1.7
link-citations: yes
bst: ../template/amnat.bst
bibliography: stomata-independence-1.bib
---

@@

```{r}
library(checkmate)
library(dplyr)
library(glue)
library(lubridate)
library(readr)
library(ropenstomata)
library(stringr)
library(tibble)

@@

# 06_connect-data ----
trimmed_data = read_rds("processed-data/trimmed-data.rds")
trimmed_phy = read_rds("processed-data/trimmed-phylogeny.rds")

@@

<!-- % Covering page info I used for New Phyt -->
<!-- % For Am Nat, moved to "template/amnat_preamble.text" -->
<!-- \begin{flushleft} -->
<!-- {\Large -->
<!-- \textbf\newline{Packing limits and developmental integration are not sufficient to explain stomatal anatomical evolution in flowering plants} -->
<!-- } -->
<!-- \newline -->
<!-- \\ -->
<!-- Christopher D. Muir\textsuperscript{1,*}, -->
<!-- Miquel \`{A}ngel Conesa\textsuperscript{2}, -->
<!-- Jeroni Galm\'{e}s\textsuperscript{2}, -->
<!-- Varsha S. Pathare\textsuperscript{3}, -->
<!-- Patricia Rivera\textsuperscript{4}, -->
<!-- Rosana López Rodríguez\textsuperscript{5}, -->
<!-- Teresa Terrazas\textsuperscript{4}, -->
<!-- Dongliang Xiong\textsuperscript{6} -->
<!-- \\ -->
<!-- \bigskip -->
<!-- \bf{1} School of Life Sciences, University of Hawaiʻi at M\=anoa, Honolulu, HI 96822, USA \\ -->
<!-- \bf{2} Research Group on Plant Biology under Mediterranean Conditions, Departament de Biologia, Universitat de les Illes Balears, Ctra. Valldemossa km 7.5, E-07122, Palma, Spain \\ -->
<!-- \bf{3} School of Biological Sciences, Washington State University, Pullman, WA 99164-4236, USA \\ -->
<!-- \bf{4} Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Apartado Postal 70‑367, 04510 Mexico City, Mexico \\ -->
<!-- \bf{5} Departamento de Sistemas y Recursos Naturales, Universidad Politécnica de Madrid, 28040 Madrid, Spain \\ -->
<!-- \bf{6} National Key Laboratory of Crop Genetic Improvement, MOA Key Laboratory of Crop Ecophysiology and Farming System in the Middle Reaches of the Yangtze River, College of Plant Science and Technology, Huazhong Agricultural University, Wuhan, Hubei 430070, China\\ -->
<!-- \bigskip -->
<!-- * cdmuir@hawaii.edu -->

@@

\textit{Word Count}: <!--4,704-->

@@

Functional and developmental constraints on phenotypic variation may cause traits to covary over millions of years and slow populations from reaching their adaptive optima. Alternatively, trait covariation may result from selective constraint if some trait combinations are generally maladaptive. Quantifying the relative contribution of functional, developmental, and selective constraints on phenotypic variation is a longstanding goal of macroevolution, but it is often difficult to distinguish different types of constraints. The stomatal anatomy of leaves with stomata on both surfaces (amphistomatous) present a unique opportunity to test the significance of functional and developmental constraints on phenotypyic evolution at broad phylogenetic scales. The key insight is that stomata on each leaf surface encounter the same functional and developmental constraints, but potentially very different selective constraints. If stomatal traits on each surface evolve differently, this implies that independent evolution is possible and functional or developmental constraints alone likely do not explain trait covariance. Packing limits and cell-size-mediated developmental integration are hypothesized to constrain variation in stomatal anatomy. We tested this by synthesizing data on stomatal density and length from amphistomatous leaves of `r n_tax3` terrestrial flowering plant taxa mostly from the literature. We estimated the covariance in divergence between stomatal traits from `r n_pair` phylogenetically independent contrasts using a robust Bayesian model. Contrary to packing limit and developmental integration hypotheses, stomatal anatomy on each surface diverged partially independently. To better understand the global variation in ecologically important traits like stomata will likely require adaptive explanations for what constrains phenotypic (co)variation.

@@

The ability for traits to evolve independently of one another is a necessary prerequisite for adaptation to complex environments [@lewontin_adaptation_1978]. If traits can evolve independently and there is sufficient genetic variation, then selection should move populations toward their multivariate phenotypic optimum. Yet divergence in one trait often covaries with other traits and covariance can persist for millions of years [@schluter_adaptive_1996]. Covariance is "a rough local measure of the strength of constraint" [@maynard_smith_developmental_1985] that can be broken down into functional, developmental, and selective constraints. Functional constraints are "limitations imposed by time, energy, or the laws of physics" [@arnold_constraints_1992]. In other words, certain trait combinations are not physically or geometrically possible. A classic example is shell coiling among invertebrate lineages in which the morphospace of possible phenotypes is constrained by hard geometrical limits [@raup_geometric_1966; @mcghee_theoretical_1999]. Within the space of possible phenotypes, developmental constraints can "bias...the production of variant phenotypes or [place] a limitation on phenotypic variability" [@maynard_smith_developmental_1985]. For example, Fibonacci phyllotaxis may arise from a packing constraint of primordia on the developing apex [@mitchison_phyllotaxis_1977; @reinhardt_law_2022; but see @niklas_role_1988 for an adaptive explanation]. Natural selection constrains phenotypic variation by preventing maladaptive forms from evolving. Selection causes trait covariation because 'missing' trait combinations are maladaptive. Understanding phenotypic constraint is challenging, but a useful starting point is determining whether phenotypic covariation can be explained by functional or developmental constraints [@mcghee_theoretical_1999; @mcghee_geometry_2007; @olson_plant_2019]. If phenotypic covariation is inconsistent with functional and developmental constraints, this provides a strong impetus to test for selective constraint.

In this study, we will address packing problems and developmental integration, specific forms of functional and developmental constraint relevant to our study system, stomatal anatomy. We introduce these concepts generally in this paragraph. Packing a number objects into a finite space is a common functional constraint on organisms. Regular geometries that appear in nature such as helices and hexagons (think DNA and honeycombs) are often optimal solutions to packing problems [@mackenzie_proving_1999; @maritan_optimal_2000]. Notice that functional constraint does not preclude selection, but the presence of a packing limit changes the range of possible phenotypes. Developmental integration is a form of developmental constraint on multivariate phenotypic evolution and we use them interchangeably in this study. Developmentally integrated traits have a "disposition for covariation" [@armbruster_integrated_2014], meaning that evolutionary divergence between lineages in one trait will be tightly associated with divergence in another trait. Allometry is a classic, albeit contested, example of developmental integration that may constrain phenotypic evolution [reviewed in @pelabon_evolution_2014]. Strong allometric covariation between traits within populations can constrain macroevolutionary divergence for long periods of time depending on the strength and direction of selection [@lande_quantitative_1979]. However, developmental integration does not necessarily hamper adaptation, and can even accelerate adaptive evolution when trait covariation is aligned with the direction of selection [@hansen_is_2003]. For example, fusion of floral parts increases their developmental integration which may increase the rate and precision of multivariate adaptation to specialist pollinators [Berg’s rule, @berg_general_1959; @berg_ecological_1960; @conner_raissa_2014; @armbruster_covariance_1999].

Biologists have studied phenotypic constraints for decades, but progress is challenging because many multivariate phenotypes are too complex or poorly understood to quantitatively distinguish functional, developmental, and selective constraints over macroevolutionary timescales. Stomatal anatomy on the leaves of flowering plants provide an exceptional opportunity because 1) there are 1000s of species to compare and 2) the main packing constraints and developmental steps are analytically tractable. This means it is possible to derive quantitative predictions and test their generality using large comparative data sets representing millions of years of evolutionary history. For this purpose, a heretofore unappreciated fact about stomata is that many leaves have stomata on both lower and upper surfaces. The packing and developmental constraints are the same for stomata on each surface, but the selective constraints may differ. Therefore, if packing and developmental constraints dominate, stomatal anatomy on each surface should diverge in concert. Failure to do so implies that independent evolution is possible and that selective constraints most likely explain at least some of the covariance between traits. Phylogenetic comparisons of stomatal anatomy provide a statistically powerful, general, and elegant way to distinguish different phenotypic constraints that would be impossible in many other traits with as much ecological significance. The next sections provide background information on stomatal anatomy, how it varies, and why functional or developmental constraints might be important. Readers not interested in this background may skip ahead to hypotheses and predictions.

@@

Stomata are microscopic pores formed by a pair of guard cells that regulate gas exchange (CO$_2$ gain and water vapor loss) on the leaf or other photosynthetic surfaces of most land plants. Stomata originated once in the history of land plants around 500 Ma, diversified rapidly in density and size, and have been maintained in most lineages except some bryophytes and aquatic plants [recently reviewed in @clark_origin_2022]. Stomata respond physiologically by opening and closing in response to light, humidity, temperature, circadian rhythm, and plant water status [@hetherington_role_2003; @lawson_guard_2020]. On this short time scale, the stomatal size, density, and distribution on a mature leaf do not change, so the maximum rate of is fixed. At somewhat longer timescales, the plant may respond to environmental cues such as light and CO$_2$ by altering stomatal anatomy in new leaves [@casson_influence_2008]. Physiological responses (aperture change) and plastic responses (new leaves with changed anatomy) may be alternative strategies for plants to acclimate to environmental change [@haworth_co-ordination_2013]. Finally, stomatal anatomy can evolve due to inherited changes in stomatal development. Plastic and genetic changes in stomatal anatomy are both ecologically important, but most studies do not use a common garden design that would tease apart their relative contribution.

We focus on anatomical variation in the density, size, and patterning of stomata on a leaf because these factors set the maximum stomatal conductance to CO$_2$ diffusing into a leaf and the amount of water that transpires from it [@sack_hydrology_2003; @franks_effect_2001; @galmes_leaf_2013; @harrison_influence_2020]. Plants typically operate below their anatomical maximum by dynamically regulating stomatal aperture. Even though operational stomatal conductance determines the realized photosynthetic rate and water-use efficiency, anatomical parameters are useful in that they set the range of stomatal function [@de_boer_optimal_2016] and are correlated with actual stomatal function under natural conditions [@murray_consistent_2020]. All else being equal, larger, more densely packed, but evenly spaced stomata increase gas exchange [@franks_maximum_2009; @dow_physiological_2014; @lehmann_effects_2015]. Smaller stomata may also be able to respond more rapidly than larger stomata, proving the ability of leaves to track short duration environmental change [@drake_smaller_2013]. Stomata are most often found only on the lower leaf surface (hypostomy), but occur on both surfaces (amphistomy) in some species [@metcalfe_anatomy_1950; @parkhurst_adaptive_1978; @mott_adaptive_1982]. Amphistomatous leaves have a second parallel pathway from the substomatal cavities through the leaf internal airspace to sites of carboxylation in the mesophyll [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984]. Thus amphistomatous leaves have lower resistance to diffusion through the airspace which increases the photosynthetic rate [@parkhurst_intercellular_1990]. If total stomatal and other conductances to CO$_2$ supply could be held constant, then an amphistomatous leaf will have a greater conductance than am otherwise identical hypostomatous leaf. The magnitude of the advantage depends on upon the resistance to diffusion through the internal airspace, which is variable among species.

The adaptive significance and ecological distribution of leaves with different stomatal anatomies is complex and there is much yet to learn. Seed plants posses a wider range of stomatal anatomies than ferns and fern allies, which are restricted to having large stomata, at low density, only on the lower surface [@de_boer_optimal_2016]. In general, trees and shrubs have greater stomatal density than herbs, but there is a of lot variation within growth forms depending on the ecological niche [@salisbury_i_1928; @kelly_plant_1995]. A commonly observed trend is that leaves from higher light environments tend to have greater stomatal density [@salisbury_i_1928; @gibson_structure-function_1996]. This may explain why, perhaps counterintuitively, plants in dry environments tend to have more stomata. Drier habitats are more open, enabling plants with higher stomatal density can photosynthesize more when water is available, but close stomata during drought [@liu_variation_2018]. Over recent human history, stomatal density has tended to decline within species as atmospheric CO$_2$ concentrations have risen [@woodward_stomatal_1987; @royer_stomatal_2001]. It is unclear whether most of this change is plastic or genetic, but the overall direction is consistent with the hypothesis that plants decrease gas exchange as CO$_2$ availability increases.

The adaptive significance of variation in stomatal ratio is uncertain, but we have some clues based on the distribution of hypo- and amphistomatous leaves. Despite the fact that amphistomy can increase photosynthesis, most leaves are hypostomatous. Amphistomy should increase photosynthesis most under saturating-light conditions where CO$_2$ supply limits photosynthesis. This may explain why amphistomatous leaves are most common in high light habitats [@salisbury_i_1928; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_associations_1998; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017], especially in herbs [@muir_light_2018]. However, the light environment alone cannot explain why hypostomatous leaves predominate in shade plants [@muir_is_2019], suggesting that we need to understand the costs of upper stomata better. Upper stomata increase the susceptibility to rust pathogens in *Populus* [@mckown_association_2014; @mckown_role_2019; @fetter_growthdefense_2021]. Amphistomy may also cause the palisade mesophyll to dry out under strong vapor pressure deficits [@buckley_how_2015]. Other hypotheses about the adaptive significance of stomatal ratio are discussed in @muir_making_2015 and @drake_two_2019.

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/concepts.pdf}
\caption{Two salient features of stomatal anatomy in flowering plants are the (a) inverse relationship between stomatal size and density and (b) the bimodal distribution of stomatal ratio. At broad phylogenetic scales, leaves with smaller stomata ($x$-axis, log-scale) tend to have greater stomatal density  ($x$-axis, log-scale), but there is a lot of variation about the overall trend indicated by the grey ellipse. Hypostomatous leaves (stomatal ratio = 0) are more common than amphistomatoues leaves, but within amphistomatous leaves, the density of stomata on each surface tends to be similar (stomatal ratio $\approx$ 0.5), which we refer to as bimodal stomatal ratio.}
\label{fig:concepts}
\end{figure}

@@

Guard cell size and spacing between stomata (the inverse of density) are developmentally intertwined because guard cells and epidermal pavement cells between stomata develop from the same meristem. Before guard cell meristemoids form via asymmetric cell division [@dow_patterning_2014], the size of guard and epidermal cells are influenced by meristematic cell volume and expansion. Evolutionary shifts in meristematic cell volume or expansion rate could cause both increased stomatal size and lower density because epidermal cells between stomata are larger [@brodribb_unified_2013]. For example, larger genomes increase meristematic cell volume [@simova_geometrical_2012], setting a lower bound on final cell volume. Although different expansion rates in guard and epidermal pavement cells can reduce the correlation in their final size, the fact that species with larger genomes tend toward having larger stomata and lower density may indicate an effect of development integration on stomatal anatomy [@beaulieu_genome_2008; @simonin_genome_2018; @roddy_scaling_2020]. Developmental integration in this case would not necessarily hinder adaptive evolution if the main axes of selection were aligned with the developmental correlation. For example, if higher maximum stomatal conductance were achieved primarily by increasing stomatal density and decreasing stomatal size as proposed by @franks_maximum_2009, then developmental integration might accelerate the response to selection compared to a case where stomatal size and density are completely independent.

@@

Both packing limits and developmental integration could contribute to inverse size-density scaling. If limits on the fraction of epidermal area occupied by stomata constrains that combinations of stomatal size and density that are evolutionarily accessible, then we predict that evolutionary divergence in stomatal size and/or density will decrease as the fraction of epidermal area occupied by stomata increases. Furthermore, if divergence slows as epidermal area occupied by stomata because of a packing limit, it should slow down the same way for both ab- and adaxial surfaces.

@@

We generated an ultrametric, bifurcating phylogeny of `r n_tax3` taxa by resolving and removing ambiguous taxonomic names, placing taxa on the GBOTB.extended mega-tree of seed plants [@smith_constructing_2018; @zanne_three_2014], and resolving polytomies using published sequence data. The complete methodology is described in the online supplement (Notes A2).

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/h3.pdf}
\caption{Evolutionary divergence slows down as epidermal space fills up. The shaded area in each facet indicates the estimate of where 95\% of the `r n_pair` phylogenetically independent contrasts fall as a function of the fraction of epidermal area allocated to stomata per surface. Each point is the absolute value of  $\Delta~\textrm{log(trait)}$ for stomatal density (left facets) or length (right facets) on the adaxial (upper facets) and abaxial (lower facets) surface. The fraction of epidermal area allocated to stomata is the average value per surface between the two taxa in each contrast. Divergence in anatomical traits is more variable when stomata occupy a smaller area, especially for adaxial stomatal density (upper left facet). See Table \ref{tab:h3output} for all parameter estimates and confidence intervals.}
\label{fig:h3}
\end{figure}

@@

```{r}
# write supporting information table
write_csv(pair_div, "ms/tableA2.csv")
```

>>>>

---
output:
  bookdown::pdf_document2:
    includes:
      before_body: ../template/amnat_doc_prefix.tex
      in_header: ../template/amnat_preamble.tex
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: no
    toc: no
  bookdown::html_document2:
    number_sections: no
    theme: readable
    toc: yes
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
  bookdown::word_document2: null
fontsize: 12pt
# linestretch: 1.0
linestretch: 1.7
link-citations: yes
bst: ../template/amnat.bst
bibliography: stomata-independence-1.bib
---

@@

```{r, message=FALSE}
library(checkmate)
library(dplyr)
library(glue)
library(lubridate)
library(readr)
library(ropenstomata)
library(stringr)
library(tibble)

@@

# 06_connect-data ----
trimmed_data = read_rds("processed-data/trimmed-data.rds")
trimmed_phy = read_rds("processed-data/trimmed-phylogeny.rds")
write_csv(trimmed_data, "dryad/trimmed-data.csv")

@@

# Write additional DRYAD files
write_csv(stomata_anatomy, "dryad/stomatal-anatomy.csv")
write_csv(stomata_anatomy_metadata, "dryad/stomatal-anatomy-metadata.csv")

invisible(file.copy("raw-data/plant-c-value.csv", "dryad/plant-c-value.csv",
          overwrite = TRUE))
```

<!-- % For Am Nat, moved to "template/amnat_preamble.text" -->
<!-- \begin{flushleft} -->
<!-- {\Large -->
<!-- \textbf\newline{Packing limits and developmental integration are not sufficient to explain stomatal anatomical evolution in flowering plants} -->
<!-- \textbf\newline{How important are functional and developmental constraints on phenotypic evolution? An empirical test with the stomatal anatomy of flowering plants} -->
<!-- } -->
<!-- \newline -->
<!-- \\ -->
<!-- Christopher D. Muir\textsuperscript{1,*}, -->
<!-- Miquel \`{A}ngel Conesa\textsuperscript{2}, -->
<!-- Jeroni Galm\'{e}s\textsuperscript{2}, -->
<!-- Varsha S. Pathare\textsuperscript{3}, -->
<!-- Patricia Rivera\textsuperscript{4}, -->
<!-- Rosana López Rodríguez\textsuperscript{5}, -->
<!-- Teresa Terrazas\textsuperscript{4}, -->
<!-- Dongliang Xiong\textsuperscript{6} -->
<!-- \\ -->
<!-- \bigskip -->
<!-- \bf{1} School of Life Sciences, University of Hawaii at M\=anoa, Honolulu, HI 96822, USA \\ -->
<!-- \bf{2} Research Group on Plant Biology under Mediterranean Conditions, Departament de Biologia, Universitat de les Illes Balears, Ctra. Valldemossa km 7.5, E-07122, Palma, Spain \\ -->
<!-- \bf{3} School of Biological Sciences, Washington State University, Pullman, WA 99164-4236, USA \\ -->
<!-- \bf{4} Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Apartado Postal 70‑367, 04510 Mexico City, Mexico \\ -->
<!-- \bf{5} Departamento de Sistemas y Recursos Naturales, Universidad Politécnica de Madrid, 28040 Madrid, Spain \\ -->
<!-- \bf{6} National Key Laboratory of Crop Genetic Improvement, MOA Key Laboratory of Crop Ecophysiology and Farming System in the Middle Reaches of the Yangtze River, College of Plant Science and Technology, Huazhong Agricultural University, Wuhan, Hubei 430070, China\\ -->
<!-- \bigskip -->
<!-- * cdmuir@hawaii.edu -->

@@

\textit{Word Count}: 4,704

@@

Functional and developmental constraints on phenotypic variation may cause traits to covary over millions of years and slow populations from reaching their adaptive optima. Alternatively, trait covariation may result from selective constraint if some trait combinations are generally maladaptive. Quantifying the relative contribution of functional, developmental, and selective constraints on phenotypic variation is a longstanding goal of macroevolution, but it is often difficult to distinguish different types of constraints. The anatomy of leaves with stomata on both surfaces (amphistomatous) present a unique opportunity to test the importance of functional and developmental constraints on phenotypyic evolution. The key insight is that stomata on each leaf surface encounter the same functional and developmental constraints, but potentially different selective constraints because of leaf asymmetry in light capture, gas exchange, and other features. Independent evolution of stomatal traits on each surface imply that functional and developmental constraints alone likely do not explain trait covariance. Packing limits on how many stomata can fit into a finite epidermis and cell-size-mediated developmental integration are hypothesized to constrain variation in stomatal anatomy. The simple geometry of the planar leaf surface and knowledge of stomatal development make it possible to derive equations for phenotypic (co)variance caused by these constraints and compare them with data. We analyzed evolutionary covariance between stomatal density and length in amphistomatous leaves from `r n_pair` phylogenetically independent contrasts using a robust Bayesian model. Stomatal anatomy on each surface diverges partially independently, meaning that packing limits and developmental integration are not sufficient to explain phenotypic (co)variation. Hence, selective constraints, which require an adaptive explanation, likely contribute to (co)variation in ecologically important traits like stomata. We show how it is possible to evaluate the contribution of different constraints by deriving expected patterns of (co)variance and testing them using similar but separate tissues, organs, or sexes.  
<!-- \newpage{} -->

@@

The ability for traits to evolve independently of one another is a necessary prerequisite for populations to ascend a fitness peak in multivariate trait space [@lewontin_adaptation_1978]. If traits can evolve independently and there is sufficient genetic variation, then selection should move populations toward their multivariate phenotypic optimum. Yet divergence in one trait often covaries with other traits and covariance can persist for millions of years [@schluter_adaptive_1996]. Covariance is "a rough local measure of the strength of constraint" [@maynard_smith_developmental_1985] that can be broken down into functional, developmental, and selective constraints. Functional constraints are "limitations imposed by time, energy, or the laws of physics" [@arnold_constraints_1992]. In other words, certain trait combinations are not physically or geometrically possible. A classic example is shell coiling among invertebrate lineages in which the morphospace of possible phenotypes is constrained by hard geometrical limits [@raup_geometric_1966; @mcghee_theoretical_1999]. Within the space of possible phenotypes, developmental constraints can "bias...the production of variant phenotypes or [place] a limitation on phenotypic variability" [@maynard_smith_developmental_1985]. For example, Fibonacci phyllotaxis may arise from a packing constraint of primordia on the developing apex [@mitchison_phyllotaxis_1977; @reinhardt_law_2022; but see @niklas_role_1988 for an adaptive explanation]. Natural selection constrains phenotypic variation by preventing maladaptive forms from evolving. Selection causes trait covariation because 'missing' trait combinations are maladaptive. Understanding phenotypic constraint is challenging, but a useful starting point is determining whether phenotypic covariation can be explained by functional or developmental constraints [@mcghee_theoretical_1999; @mcghee_geometry_2007; @olson_plant_2019]. If phenotypic covariation is inconsistent with functional and developmental constraints, this provides a strong impetus to test for selective constraint. Selective constraint does not mean that populations do not respond to selection, but rather that limits on phenotypic variation arise because selection never favors certain trait values.

In this study, we will address packing problems and developmental integration, specific forms of functional and developmental constraint relevant to our study system, stomatal anatomy. We introduce these concepts generally in this paragraph. Packing a number objects into a finite space is a common functional constraint on organisms. Regular geometries that appear in nature such as helices and hexagons (think DNA and honeycombs) are often optimal solutions to packing problems [@mackenzie_proving_1999; @maritan_optimal_2000]. Notice that functional constraint does not preclude selection, but the presence of a packing limit changes the range of possible phenotypes. Developmental integration is a form of developmental constraint on multivariate phenotypic evolution and we use these terms interchangeably in this study. Developmentally integrated traits have a "disposition for covariation" [@armbruster_integrated_2014], meaning that evolutionary divergence between lineages in one trait will be tightly associated with divergence in another trait. Allometry is a classic, albeit contested, example of developmental integration that may constrain phenotypic evolution [reviewed in @pelabon_evolution_2014]. Strong allometric covariation between traits within populations can constrain macroevolutionary divergence for long periods of time depending on the strength and direction of selection [@lande_quantitative_1979]. However, developmental integration does not necessarily hamper adaptation, and can even accelerate adaptive evolution when trait covariation is aligned with the direction of selection [@hansen_is_2003]. For example, fusion of floral parts increases their developmental integration which may increase the rate and precision of multivariate adaptation to specialist pollinators [Berg’s rule, @berg_general_1959; @berg_ecological_1960; @conner_raissa_2014; @armbruster_covariance_1999].

Biologists have studied phenotypic constraints for decades, but progress is challenging because many multivariate phenotypes are too complex or too poorly understood to quantitatively distinguish functional, developmental, and selective constraints over macroevolutionary timescales. Stomatal anatomy on the leaves of flowering plants provide an exceptional opportunity because 1) there are 1000s of species to compare and 2) the main packing constraints and developmental steps are analytically tractable. This means it is possible to derive quantitative predictions and test their generality using large comparative data sets representing millions of years of evolutionary history. For this purpose, a heretofore unappreciated fact about stomata is that many leaves have stomata on both lower and upper surfaces. The packing and developmental constraints are the same for stomata on each surface, but the selective constraints may differ. Therefore, if packing and developmental constraints dominate, stomatal anatomy on each surface should diverge in concert. Failure to do so implies that independent evolution is possible and that selective constraints most likely explain at least some of the covariance between traits. Phylogenetic comparisons of stomatal anatomy provide a statistically powerful, general, and elegant way to distinguish different phenotypic constraints that would be impossible in many other traits with as much ecological significance. The next sections provide background information on stomatal anatomy, how it varies, and why functional or developmental constraints might be important.

@@

Stomata are microscopic pores formed by a pair of guard cells that regulate gas exchange (CO$_2$ gain and water vapor loss) on the leaves or other photosynthetic surfaces of most land plants. Stomata originated once in the history of land plants around 500 Ma, diversified rapidly in density and size, and have been maintained in most lineages except some bryophytes and aquatic plants [recently reviewed in @clark_origin_2022]. Stomata respond physiologically by opening and closing in response to light, humidity, temperature, circadian rhythm, and plant water status [@hetherington_role_2003; @lawson_guard_2020]. The stomatal size, density, and distribution on a mature leaf do not change, so the maximum rate of gas exchange is fixed. However, the plant may respond plastically to environmental cues such as light and CO$_2$ by altering stomatal anatomy in new leaves [@casson_influence_2008]. Physiological responses (aperture change) and plastic responses (new leaves with changed anatomy) may be alternative strategies for plants to acclimate to environmental change [@haworth_co-ordination_2013]. Finally, stomatal anatomy can evolve due to inherited changes in stomatal development. Plastic and genetic changes in stomatal anatomy are both ecologically important, but most studies do not use a common garden design that would tease apart their relative contribution.

We focus on anatomical variation in the density, size, and patterning of stomata on a leaf because these factors set the maximum stomatal conductance to CO$_2$ diffusing into a leaf and the amount of water that transpires from it [@sack_hydrology_2003; @franks_effect_2001; @galmes_leaf_2013; @harrison_influence_2020]. Plants typically operate below their anatomical maximum by dynamically regulating stomatal aperture. Even though operational stomatal conductance determines the realized photosynthetic rate and water-use efficiency, anatomical parameters are useful in that they set the range of stomatal function [@de_boer_optimal_2016] and are correlated with actual stomatal function under natural conditions [@murray_consistent_2020]. All else being equal, larger, more densely packed, but evenly spaced stomata increase gas exchange [@franks_maximum_2009; @dow_physiological_2014; @lehmann_effects_2015]. Smaller stomata may also be able to respond more rapidly than larger stomata, proving the ability of leaves to track short duration environmental change [@drake_smaller_2013]. Stomata are most often found only on the lower leaf surface (hypostomy), but occur on both surfaces (amphistomy) in some species [@metcalfe_anatomy_1950; @parkhurst_adaptive_1978; @mott_adaptive_1982]. Amphistomatous leaves have a second parallel pathway from the substomatal cavities through the leaf internal airspace to sites of carboxylation in the mesophyll [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984]. Thus amphistomatous leaves have lower resistance to diffusion through the airspace which increases the photosynthetic rate [@parkhurst_intercellular_1990]. If total stomatal and other conductances to CO$_2$ supply could be held constant, then an amphistomatous leaf will have a greater conductance than an otherwise identical hypostomatous leaf. The magnitude of the advantage depends on the resistance to diffusion through the internal airspace, which is variable among species.

The adaptive significance and ecological distribution of leaves with different stomatal anatomies is complex and there is much yet to learn. Seed plants posses a wider range of stomatal anatomies than ferns and fern allies, which are restricted to having large stomata, at low density, only on the lower surface [@de_boer_optimal_2016]. In general, trees and shrubs have greater stomatal density than herbs, but there is a of lot variation within growth forms depending on the ecological niche [@salisbury_i_1928; @kelly_plant_1995]. A commonly observed trend is that leaves from higher light environments tend to have greater stomatal density [@salisbury_i_1928; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_associations_1998; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017]. This may explain why, perhaps counterintuitively, plants in dry environments tend to have more stomata. Drier habitats are more open, enabling plants with higher stomatal density to photosynthesize more when water is available, but close stomata during drought [@liu_variation_2018]. Over recent human history, stomatal density has tended to decline within species as atmospheric CO$_2$ concentrations have risen [@woodward_stomatal_1987; @royer_stomatal_2001]. It is unclear whether most of this change is plastic or genetic, but the overall direction is consistent with the hypothesis that plants decrease gas exchange as CO$_2$ availability increases.

Stomatal ratio is a continuous measure of how stomata are deployed across leaf surfaces. The adaptive significance of variation in stomatal ratio is uncertain, but we have some clues based on the distribution of hypo- and amphistomatous leaves. Despite the fact that amphistomy can increase photosynthesis, most leaves are hypostomatous. Amphistomy should increase photosynthesis most under saturating-light conditions where CO$_2$ supply limits photosynthesis. However, the light environment alone cannot explain why hypostomatous leaves predominate in shade plants [@muir_is_2019], suggesting that we need to understand the costs of upper stomata better. One factor may be increased vulnerability to pathogens. For example, upper stomata increase susceptibility to rust pathogens in *Populus* [@mckown_association_2014; @mckown_role_2019; @fetter_growthdefense_2021]. Amphistomy may also cause the palisade mesophyll to dry out under strong vapor pressure deficits [@buckley_how_2015]. Other hypotheses about the adaptive significance of stomatal ratio are discussed in @muir_making_2015 and @drake_two_2019.

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/concepts.pdf}
\caption{Two salient features of stomatal anatomy in flowering plants are the (a) inverse relationship between stomatal size and density and (b) the bimodal distribution of stomatal ratio. At broad phylogenetic scales, species with smaller stomata on their leaves ($x$-axis, log-scale) tend to have greater stomatal density  ($x$-axis, log-scale), but there is a lot of variation about the overall trend indicated by the grey ellipse. Hypostomatous leaves (stomatal ratio = 0) are more common than amphistomatoues leaves, but within amphistomatous leaves, the density of stomata on each surface tends to be similar (stomatal ratio $\approx$ 0.5), which we refer to as bimodal stomatal ratio.}
\label{fig:concepts}
\end{figure}

@@

Guard cell size and spacing between stomata (the inverse of density) are developmentally intertwined because guard cells and epidermal pavement cells between stomata develop from the same meristem. Before guard cell meristemoids form via asymmetric cell division [@dow_patterning_2014], the size of guard and epidermal cells are influenced by meristematic cell volume and expansion. Evolutionary shifts in meristematic cell volume or expansion rate could cause both increased stomatal size and lower density because epidermal cells between stomata are larger [@brodribb_unified_2013]. For example, larger genomes increase meristematic cell volume [@simova_geometrical_2012], which sets a lower bound on final cell volume. Although different expansion rates in guard and epidermal pavement cells can reduce the correlation in their final size, the fact that species with larger genomes tend toward having larger stomata and lower density may indicate an effect of development integration on stomatal anatomy [@beaulieu_genome_2008; @simonin_genome_2018; @roddy_scaling_2020]. Developmental integration in this case would not necessarily hinder adaptive evolution if the main axes of selection were aligned with the developmental correlation. For example, if higher maximum stomatal conductance were achieved primarily by increasing stomatal density and decreasing stomatal size as proposed by @franks_maximum_2009, then developmental integration might accelerate the response to selection compared to a case where stomatal size and density are completely independent.

@@

Both packing limits and developmental integration could contribute to inverse size-density scaling. If limits on the fraction of epidermal area occupied by stomata constrains the combinations of stomatal size and density that are evolutionarily accessible, then we predict that evolutionary divergence in stomatal size and/or density will decrease as the fraction of epidermal area occupied by stomata increases. Furthermore, if divergence slows as epidermal area occupied by stomata because of a packing limit, it should slow down the same way for both ab- and adaxial surfaces.

@@

We generated an ultrametric, bifurcating phylogeny of `r n_tax3` taxa by resolving and removing ambiguous taxonomic names, placing taxa on the GBOTB.extended mega-tree of seed plants [@smith_constructing_2018; @zanne_three_2014], and resolving polytomies using published sequence data. Divergence times in this phylogeny are based on extensive fossil calibration [see @smith_constructing_2018; @magallon_metacalibrated_2015 for further detail]. The complete methodology is described in the online supplement (Notes A2).

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/h3.pdf}
\caption{Evolutionary divergence slows down as epidermal space fills up. This pattern is consistent with functional constraints (packing limits) constraining evolution, but only when stomata occupy a large fraction of epidermal area. The shaded area in each facet indicates the estimate of where 95\% of the `r n_pair` phylogenetically independent contrasts fall as a function of the fraction of epidermal area allocated to stomata per surface. Each point is the absolute value of  $\Delta~\textrm{log(trait)}$ for stomatal density (left facets) or length (right facets) on the adaxial (upper facets) and abaxial (lower facets) surface. The fraction of epidermal area allocated to stomata is the average value per surface between the two taxa in each contrast. Divergence in anatomical traits is more variable when stomata occupy a smaller area, especially for adaxial stomatal density (upper left facet). See Table \ref{tab:h3output} for all parameter estimates and confidence intervals.}
\label{fig:h3}
\end{figure}

@@

```{r}
# write supporting information table
write_csv(pair_div, "ms/tableA1.csv")
# write data set for DRYAD
write_csv(pair_div, "dryad/phylogenetic-contrasts.csv")
```
